Lazy Linearity for a Core Functional Language

### 2025-10-18 popl26.tex

\ROUNDTWO{The reader may wonder whether the complexities of our
  system hinted above are a better alternative to simply specializing
  the optimisation passes to produce code that passes a more naive
  linearity check. This avenue was already explored during the
  implementation of Linear
  Haskell in
  GHC\footnote{https://gitlab.haskell.org/ghc/ghc/-/issues/22123},
 exhibiting the prohibitive cost of too often restricting
 optimisations and producing less efficient code. Our work adheres to
 the principle that the type system bends to the optimisations, not
 the other way around.}

@@

\paragraph{Linear Core in the Glasgow Haskell Compiler.}
% Linear Core is suitable as the intermediate language of an optimising compiler
% for a linear and lazy language such as Haskell Core, in that optimising
% transformations in Linear Core preserve types \emph{and} linearity, since Linear
% Core understands semantic linearity in the presence of laziness, unlike
% Core's current type system under which optimisations currently violate
% linearity.
%
Integrating Linear Core in the Glasgow Haskell Compiler is one of the ultimate
goals of our work. Core's current type system ignores linearity due to
its limitation in understanding semantic linearity, and our work fills this gap
and would allow Core to be linearly typed all throughout.
%
% A linearly typed Core that preserves linearity throughout the optimisation
% pipeline of GHC both validates the correctness of the compiler, which is
% already achieved to a great extent by preserving (non-linear) types, and
% informs optimisations, allowing the compiler to generate more performant programs.
%
Implementing Linear Core in GHC is a challenging endeavour, since we must
account for all other Core features (e.g. strict constructor fields), propagate
new or modified data throughout the entire pipeline, and accept more
optimisations. Despite our initiative in this direction\footnote{https://gitlab.haskell.org/ghc/ghc/-/issues/23218},
we leave this as future work.

>>>>

\ROUNDTWO{The reader may wonder whether the complexities of our
  system hinted above are a better alternative to simply specializing
  the optimisation passes to produce code that passes a more naive
  linearity check. This avenue was already explored during the
  implementation of Linear
  Haskell in GHC,
 exhibiting the prohibitive cost of too often restricting
 optimisations and producing less efficient code. Our work adheres to
 the principle that the type system bends to the optimisations, not
 the other way around\footnote{\url{https://gitlab.haskell.org/ghc/ghc/-/blob/c85c845dc5ad539bf28f1b8c5c1dbb349e3f3d25/compiler/GHC/Core/Lint.hs\#L3225-3235}}.}

@@

\paragraph{Linear Core in the Glasgow Haskell Compiler.}
% Linear Core is suitable as the intermediate language of an optimising compiler
% for a linear and lazy language such as Haskell Core, in that optimising
% transformations in Linear Core preserve types \emph{and} linearity, since Linear
% Core understands semantic linearity in the presence of laziness, unlike
% Core's current type system under which optimisations currently violate
% linearity.
%
Integrating Linear Core in the Glasgow Haskell Compiler is one of the ultimate
goals of our work. Core's current type system ignores linearity due to
its limitation in understanding semantic linearity, and our work fills this gap
and would allow Core to be linearly typed all throughout.
%
% A linearly typed Core that preserves linearity throughout the optimisation
% pipeline of GHC both validates the correctness of the compiler, which is
% already achieved to a great extent by preserving (non-linear) types, and
% informs optimisations, allowing the compiler to generate more performant programs.
%
Implementing Linear Core in GHC is a challenging endeavour, since we must
account for all other Core features (e.g. strict constructor fields), propagate
new or modified data throughout the entire pipeline, and accept more
optimisations. Despite our initiative in this direction\footnote{\url{https://gitlab.haskell.org/ghc/ghc/-/merge_requests/10310}},
we leave this as future work.
